Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-29 manu-draft.qmd
methods revision done

Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed. Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the reaminder of the watershed).

Subtidal included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] were classified as emergent tidal wetlands. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.*  Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).

@@

Opportunity areas for restoration and recommended coverage targets for habitat types were identifed by integrating multiple datasets available for the region.  Habitat status and historical trends were first quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.  Historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.  Finally, remaining opportunities for restoration that integrated results from the first two analyses were identified based on current status of habitat coverage in the watershed.  This included combining current coverages with existing protected areas and considering their location within the three strata (sub-, inter-, supratidal) and areas anticipated to be impacted by sea-level rise and development near coastal areas.  The final results included spatially-explicit maps that identified opportunity areas and restoration potential for remaining native habitats in the watershed.  Short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats were informed by current status, past trends, and historical restoration efforts in the region.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19]. Historical datasets for subtidal habitats begin in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) geospatial database [@SWFWMD18]. Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands from non-forested wetlands can be distinguished with these data. Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications. Native upland habitats were also combined in one classification. Historical estimates for these habitat coverages were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.

To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates. These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.

Current and historical coverage estimates were obtained for all habitat types, including estimates of developed lands, for each year of available data spanning approximately three decades at two-year time steps.  In addition, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types.  This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).  The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years.

@@

Past restoration projects were quantified for each of the major habitat types to inform expectations for setting short-term goals and long-term targets. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database. Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, lead partner, size (area or length), and latitude and longitude. Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals. Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.

@@

Spatially-explicit estimates of the remaining opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.  The two main processes included binning existing datasets into relevant categories and overlaying multiple datasets to identify restoration opportunities.  Opportunity areas were defined as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets. The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related both to land use change and sea-level rise.

The Land Use Land Cover 2017 dataset from SWFWMD was used as the baseline for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata.

 All FLUCCS classification codes were placed into one of three categories. First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands). Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography. Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.

Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework. The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070. The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea-level rise. The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed. Floodplain corridors provide vital watershed functions including fish and wildlife habitat and migratory pathways, floodwater attenuation and storage, erosion control, and delivery of complex organic matter to the estuarine food web. Finally, the upland stratum encompasses those areas outside of the coastal and river floodplain strata, including native upland habitats as well as hydrologically isolated wetlands. These habitats provide important aquifer recharge and wildlife habitat functions.

Layers for existing public lands and parcels targeted for acquisition were compiled by combining data from the Florida Natural Areas Inventory (Table 1), consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data.

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soils distributions can be used to generally represent historical habitat distribution, and can be used to provide generalized restoration guidelines (e.g., tidal wetlands, freshwater wetlands, and native uplands). @Ries14 created a soils suitability analysis for wetland mitigation and restoration sing data from the USDA Web Soil Survey (Table 1) and classified all soils in the Tampa Bay soils into one of three categories (xeric, mesic, and hydric). The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.

A distinction was made between tidal and freshwater wetland restoration potential by intersecting the combined mesic and hydric soils polygons with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were classified as having tidal wetland restoration potential, while mesic or hydric soils occurring above the 5-foot contour were classified as having freshwater wetland restoration potential.

Within the Tampa Bay region, lands adjacent to waters with average salinity values greater than 18 psu are considered most appropriate for higher salinity mangrove/salt barren restoration, while lands adjacent to waters with average salinity values less than 18 psu are considered most appropriate for lower salinity salt marsh (*Juncus* spp.) restoration [@Eleuterius79]. To estimate the relative restoration potential of mangrove/salt barrens and salt marshes, a regional long-term water quality data set was used to create salinity isohalines, which was then binned into two salinity categories: greater and less than an annual mean of 18 psu (Table 1).

All acquired and created GIS layers (Table 1) were converted to 10m x 10m raster data sets and delimited using a previously established watershed boundary. The reservation, restorable, conservation, acquisition, soils, and salinity isohaline layers were then merged to create a comprehensive set of maps and matrices of all possible combinations of opportunity areas. Within an ArcGIS Pro 2.x GIS environment, the SWFWMD 2017 FLUCCS and 2018 submerged data was classified and extracted to the study area into described categories. The dataset was then spatially assigned to existing and proposed restoration areas, reservation areas, soil type, and salinity level.

>>>>

The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed. Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the remainder of the watershed).

Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habiats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour area was considered the most likely to be impacted by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@tbep0519].

@@

Opportunity areas for restoration and recommended coverage targets for habitat types were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners. The first two steps were used to identify short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats. Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise. As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19]. Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) geospatial database [@SWFWMD18]. Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data. Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications. Native upland habitats were also combined in one classification. Historical estimates for these habitat coverages were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.

To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates. These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines, only available as current estimates with no information on historical trends.

Current and historical coverage estimates were obtained for all habitat types, including estimates of developed lands, for each year of available data spanning approximately three decades at two-year time steps. In addition, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types. This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats). The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].

@@

Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single database. Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, size (area or length), and location (latitude and longitude). Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals. Living shoreline projects, including seawall enhancements and oyster reef installations, were inventoried separately.

@@

Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow. The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify restoration opportunities. Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain defined goals and targets. The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related to anticipated land use change and sea-level rise.

The Land Use Land Cover 2017 dataset from SWFWMD was used for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata. All FLUCCS classification codes were placed into one of three categories. First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands). Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration. Third, existing development included developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration.

After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas. Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed. The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum. This area extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519]. Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.

The existing and proposed native and restorable areas were those that occured in public lands and parcels currently acquired or targeted for acquisition. To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory, consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data. This workflow created the existing conservation and proposed (i.e., planned to be acquired) conservation layers in @fig-gisworkflow . Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation native.

All opportunity areas identified as restorable included reservation restorable (in the coastal stratum), existing conservation restorable, and proposed conservation restorable. These layers were further grouped into their restoration potential as discrete habitat types that could be the goal of future restoration projects. The restoration potential was defined using the coastal stratum, soil types, and salinities of adjacent waters.

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential. @Ries14 created a soils suitability analysis for wetland mitigation and restoration and classified all soils in the Tampa Bay watershed as xeric, mesic, or hydric. The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.

A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands. A final restoration potential category was created for low salinity salt marshes (*Juncus* spp.) in the coastal stratum. These included restorable lands with mesic or hydric soils and also lands adjacent to waters with average salinity values less than 18 psu [@Eleuterius79]. The salinity values were identified using salinity isohalines created from a long-term water quality data set provided by the Environmental Protection Commission of Hillsborough County.

Two distinct mapping products were created from the above analysis. The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable. The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes). All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging tools from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18]. The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].
